\begin{abstract}
Conditional Flow Matching (CFM) enables simulation-free training of continuous normalizing flows by regressing a neural velocity field against a target conditional vector field. However, the single-sample conditional target exhibits high gradient variance, which degrades sample quality and slows convergence. Three independent lines of work address this variance through different axes: \emph{coupling} (replacing independent noise--data pairing with minibatch optimal transport), \emph{objective} (approximating the marginal vector field via importance-weighted multi-reference targets), and \emph{timestep estimation} (antithetic temporal sampling with consistency regularization). Whether these techniques compose additively or interfere remains unexplored.

We investigate this question through two complementary studies. \textbf{Study~1} compares four flow-matching variants (OT-CFM, VP-CFM, Target-CFM, SB-CFM) on 2D synthetic distributions, confirming that OT-type paths produce straighter trajectories and better low-budget robustness, while Schr\"odinger Bridge paths achieve the best endpoint quality. \textbf{Study~2} crosses the three variance-reduction axes in a full $2^3$ factorial ablation on CIFAR-10 (8 cells, U-Net backbone, 80k training steps). Only one cell converged: BatchOT + CFM + Uniform (cell\_100), achieving FID~18.4. The remaining seven cells diverged to NaN or exhibited severely unstable training.

These negative results are informative: they reveal that StableVM's importance weights scale catastrophically with data dimension ($d{=}3072$), that TPC's consistency penalty introduces gradient explosions at complementary timesteps, and that BatchOT coupling is the single most impactful variance-reduction technique. The three axes are not additive---combining them is destructive in high dimensions.
\end{abstract}
